Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-03 supplement.tex
diversity stats and relate inference

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

@@

@@

All analyses presented in this work focus on biallelic single nucleotide
polymorphisms within the 1000 Genomes Phase 3 strict mask.

For the moments-LD analysis, we focused on intergenic locations because these
appear less affected by natural selection compared to both synonymous and
nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
had less than 100 contiguous base pairs.

@@

\aprcomment{$D^2$, $D_z$, $\pi_2$, and $H$. Definitions, notation,
indexing\ldots}

@@

@@

\section{Validations using simulatinos from inferred demographic models}

@@

>>>>

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}

@@

\newcommand{\E}{\mathbb{E}}

@@

All analyses presented in this work focus on biallelic single nucleotide
polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the moments-LD
analysis, we focused on intergenic locations because these appear less affected
by natural selection compared to both synonymous and nonsynonymous variation
\citep{Ragsdale2018-dd}. To enable comparison with Neanderthal DNA, we excluded
regions for which the Vindija Neanderthal sample had less than 100 contiguous
base pairs.

@@

We used multi-population linkage disequilibrium (LD) and pairwise diverisity
statistics to fit demographic models to data. These statistics, introduced and
described in detail in \citet{Ragsdale2019-nt}, are the multi-population analog
of the classic LD statistics first described and computed by
\citet{Hill1968-vu,Ohta1971-yd}.

Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
allels $B$ and $b$ at the right locus, the standard covariance measure of LD is
$D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
haplotypes in a population (and thus the probability of drawing an $AB$
haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
the expectation of $D^2$ using a system of equations that includes
$\E[Dz] = \E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] =
\E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$ and $p_B$ are the frequencies of
$A$ and $B$ at the left and right loci, respectively. This system also requires
the expected pairwise diversity (or expected heterozygosity, assuming random
mating), denoted $\E[H] = \E[2p_A(1-p_A)] =
\E[2p_B(1-pB)]$, assuming equal mutation rates at the two loci.

\citet{Ragsdale2019-nt} showed how to compute the analogous multi-population
set of LD statistics, and we refer readers there for details on their
definitions, computation, and interpretations. In short, we obtain expectations
of $D^2$ in each population, the cross-population product $D_iD_j$ (where $i$
and $j$ index populations), as well as those additional statistics $Dz$ and
$\pi_2$ taken over different combinations of population indexing. That is
$Dz_{i, j, k} = D_i(1-2p_{A,j})(1-2p_{B,k})$ and
$\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
normalized by $\pi_2$ in a reference population (throughout, we use the Mende
$\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
$\E[H_{i,j}] = \E[p_i(1-p_j) + p_j(1-p_i)]$ were also normalized by $H$ in the
Mende, so that all pairwise diversity measures are relative to the reference
population.

$\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
archaic admixture \citep{Ragsdale2019-nt}, and this statistic is closely
related to $S^*$ statistics used to scan for introgressed haplotypes
\citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
widely used in demographic inference involving ancient DNA and many samples, as
$f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
$H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
\aprcomment{cite}. Therefore, the set of statistics used here encompass
multiple features of genetic data that have been used to infer models of
archaic admixture and population structure involving many populations.

@@

We used Relate version 1.0.16 \citep{Speidel2019-nj} to reconstruct genome-wide
gene genealogies using a combined set of 1000 Genomes and African Diversity
Reference Panel datasets, retaining all AFR-labeled populations and GBR, CEU,
PJL, and CHB from the 1000 Genomes panel and the Nama, Gumuz, Oromo, and Amhara
from ADRP. We used all autosomes and applied the 1000 Genomes Phase 3 strict
mask, we used an ancestral sequence determined by a 6-primate alignment
(human\_ancestor\_GRCh37\_e59), and we used the HapMap II combined
recombination map, all in GRCh37 coordinates. We assumed a generation time of
29 years, a mutation rate of 1.25e-8, and followed the standard pipeline
described in the Relate documentation.

From the reconstructed gene genealogies, we computed coalescence rates within
and between populations using Relate’s function RelateCoalescentRate. This
allows for an estimate of the instantaneous inverse coalescence rate (IIRC) for
samples drawn within each population (the inverse of which is often interpreted
as the effective population size history), and the relative cross coalescence
rates between pairs of populations (which are commonly used to estimate
divergence times).

Following \citet{Speidel2019-nj} we also identified “deep branches” within gene
trees. Such a “deep branch” is a branch within a marginal tree that has its
upper end (or node) extending to more than 1 million years in age, and we
partitioned deep branches based on their lower end ages into bins between 0 and
1Ma. Such branches can be categorized by their association with Neanderthals by
comparing mutations that fall upon such a branch to the allele found in a
Neanderthal genome sequence. For this, we used the published high-coverage
Vindija Neanderthal \citep{Prufer2017-kk}. Again following the analysis in
\citet{Speidel2019-nj}, if one or more mutations on a deep branch are shared
with the Neanderthal sequence, the branch is inferred to have passed through
the Neanderthal lineage.

A deep branch is assigned to a contemporary population if at least one sample
from that population has ancestry that passes through that branch.
\citet{Speidel2019-nj} show that deep branches with lower ends more recent than
the Neanderthal introgression event are enriched for Neanderthal-matching
alleles in Eurasian populations, while a large majority of deep branches in
1000 Genomes West African populations do not match either the Neanderthal or
Denisovan samples. This observation was taken as further evidence for deep
population structure or archaic admixture in West African populations from an
unidentified hominin unrelated to the Neanderthal/Denisovan complex.

@@

\section{Validations using simulations from inferred demographic models}

@@

\section{Supplementary results}

\subsection{Conditional SFS}
